Black-swan events in animal populations

### 2016-07-13 text/ms_pnas.tex
Copyediting

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Black swans---statistically improbable events
with profound consequences---happen
more often than expected in financial, social, and natural systems. Our work
demonstrates the rare but systematic presence of black-swan events in animal populations
around the world (mostly birds, mammals, insects, and fishes). These events are
predominantly downward, implying that unexpected population crashes occur more
frequently than increases. Black-swan events are not driven by life history (e.g.\ lifespan),
but by external causes such as extreme winters and disease. Ignoring the presence of
downward black swans may severely underestimate extinction risk in animal populations,
particularly under a changing climate where such extreme events are expected to increase
in frequency and magnitude.}

@@

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied
to a large number of time series can yield insights into the frequency, strength,
and correlates of black-swan events.
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to time series
have not accounted for an underlying population dynamics model
\cite{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a chosen threshold of decline
\cite{young1994, gerber2001, reed2003, fey2015}.
But this approach
does not distinguish events from
expected dynamics.
For instance, the range of natural variability
can differ dramatically across taxa.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

We systematically evaluated the presence of black-swan events in hundreds of
animal populations.
We find strong evidence for their occurrence in animal populations. Black-swan
population crashes are substantially more frequent
than black-swan population
increases and are usually driven by external events such as weather and
disease.
Thus our analysis provides strong evidence
for downward-skewed heavy-tailed events in abundance time series of higher
taxa, and ignoring these events will tend to underestimate the risk of
population declines.
Next we consider
(1) the possible mechanisms underlying black-swan events,
(2) caveats to our findings,
(3) the consequences of our findings for making risk forecasts, and
(4) for managing natural resources.

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy-tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full ess %%% generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \cite{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
(e.g.~the $t$ distribution compared to a mixture
of two normal distributions \cite{ward2007}).
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution.
Furthermore, if the degrees of freedom parameter ($\nu$)
is fixed,
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 5$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4A)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 5$ allows for an event
three standard deviations below the mean to happen
% 1/(1-pt(3, 5))
once every 66 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Black swans---statistically improbable events
with profound consequences---happen
more often than expected in financial, social, and natural systems. Our work
demonstrates the rare but systematic presence of black-swan events in animal populations
around the world (mostly birds, mammals, and insects). These events are
predominantly downward, implying that unexpected population crashes occur more
frequently than increases. Black-swan events are not driven by life history (e.g.\ lifespan),
but by external causes such as extreme winters and disease. Ignoring the presence of
downward black swans may severely underestimate extinction risk in animal populations,
particularly under a changing climate where such extreme events are expected to increase
in frequency and magnitude.}

@@

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied
to a large number of time series can yield insights into the frequency, strength,
and correlates of such events.
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to time series
have not accounted for an underlying population dynamics model
\cite{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a chosen threshold of decline
\cite{young1994, gerber2001, reed2003, fey2015}.
But this approach
does not distinguish events from
expected dynamics.
For instance, the range of natural variability
can differ dramatically across taxa.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the extent to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

We systematically evaluated the prevalence of black-swan events in hundreds of
animal populations.
We find strong evidence for their occurrence in animal populations. Black-swan
population crashes are substantially more frequent
than black-swan population
increases and are usually driven by external events such as weather and
disease.
Thus our analysis provides strong evidence
for downward-skewed heavy-tailed events in abundance time series of higher
taxa, and ignoring these events will tend to underestimate the risk of
population declines.
Next we consider
(1) the possible mechanisms underlying black-swan events,
(2) caveats to our findings,
(3) the consequences of our findings for making risk forecasts, and
(4) for managing natural resources.

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy-tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full ess %%% generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \cite{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
(e.g.~the $t$ distribution compared to a mixture
of two normal distributions \cite{ward2007}).
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution.
Furthermore, if the degrees of freedom parameter ($\nu$)
is fixed,
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 5$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4A)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and compared to the normal distribution, $\nu  = 5$ allows for an event
three standard deviations below the mean to happen
% 1/(1-pt(3, 5))
once every 66 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.
